---
title: ""
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: no
    toc: yes
    toc_float: yes
  pdf_document:
    number_sections: no
    toc: yes
---

# Maturity ogive for the southern hake stock  {.tabset}

**Below the objective and a theoretical explanation of the model are reported. However the details of both can be find through the document tabs which explain the analysis step by step.**

**Objective**: A combined maturity ogive (maturity proportions-at-length) for the southern hake stock estimated through the data derived from both institutes (laboratories), IPMA (Instituto Português do Mar e da Atmosfera) and IEO (Instituto Español de Oceanografía). 

## Model (theoretical explanation) {.tabset}



Maturity proportions-at-length have been estimated by bayesian regression models using the integrated nested Laplace approximation (INLA) (Rue et al., 2009) approach in the R-INLA software (https://www.r-inla.org/). 

For estimating a combined maturity ogive for both laboratories a bivariate model has been required (Zuur and Ieno, 2018, additional details in Paradinas et al., 2017 and Izquierdo et al., 2021). The bivariate response variable is defined as follows.

$y_i^{IEO} \sim Bernoulli(\pi_i^{IEO})$, $i=1, \dots, N^{IEO}$; being $N^{IEO}$ the number of individuals measured by IEO.
$y_j^{IPMA} \sim Bernoulli(\pi_j^{IPMA})$, $j=1, \dots, N^{IPMA}$; being $N^{IPMA}$ the number of individuals measured by IPMA.


The covariables (explanatory variables) are the length and the year. The length variable is introduced linear. On the other hand, the year covariable is introduced differently depending on the aim: a standard year combined maturity ogive (Approach 1)  or a combined maturity ogive by year (Approach 2). 

**Approach 1**

The year variability is taken into account through the random effect $a_i, a_j \sim N(0, \sigma_{year}^2)$, $i=1, \dots, N^{IEO}$, $j=1, \dots, N^{IPMA}$. Note that $\sigma_{year}^2$ parameter is common for IEO and IPMA response variables. 


$$Logit(\pi_i^{IEO})=\ln(\pi_i^{IEO}/(1- \pi_i^{IEO}))=\beta_0+\beta_1 \times(l^{IEO}(i))+ a_i + \epsilon_i$$
$$Logit(\pi_j^{IPMA})=\ln(\pi_j^{IPMA}/(1- \pi_j^{IPMA}))=\beta_0+\beta_1 \times(l^{IPMA}(j))+ a_j + \epsilon_i$$

$l^{IEO}(i)$ assigns to each individual of IEO its corresponding length. The same for $l^{IPMA}(j)$.
$\epsilon_i, \epsilon_j \sim N(0,\sigma_{\epsilon}^2)$; $a_i,a_j \sim N(0, \sigma_{year}^2)$.


**Approach 2**

The year is included in the model as a factor covariable.


$$Logit(\pi_i^{IEO})=\ln(\pi_i^{IEO}/(1- \pi_i^{IEO}))=\beta_0+\beta_1 \times(l^{IEO}(i))+ year_i + \epsilon_i$$
$$Logit(\pi_j^{IPMA})=\ln(\pi_j^{IPMA}/(1- \pi_j^{IPMA}))=\beta_0+\beta_1 \times(l^{IPMA}(j))+ year_j + \epsilon_i$$

$l^{IEO}(i)$ assigns to each individual of IEO its corresponding length. The same for $l^{IPMA}(j)$. $year_i$, $year_j$ is a categorical covariate allowing for a different mean value per year.
$\epsilon_i, \epsilon_j \sim N(0,\sigma_{\epsilon}^2)$.


**References**

- Izquierdo, F., Paradinas, I., Cerviño, S., Conesa, D., Alonso-Fernández, A., Velasco, F., ... & Pennino, M. G. (2021). Spatio-temporal assessment of the European hake (Merluccius merluccius) recruits in the northern Iberian Peninsula. Frontiers in Marine Science, 8, 1.
- Paradinas, I., Conesa, D., Lopez-Quilez, A., & Bellido, J. M. (2017). Spatio-temporal model structures with shared components for semi-continuous species distribution modelling. Spatial Statistics, 22, 434-450.
- Rue, H., Martino, S., and Chopin, N. (2009). Approximate Bayesian inference for latent Gaussian models by using integrated nested Laplace approximations. J. R. Stat. Soc. B. 71, 319–392. doi: 10.1111/j.1467-9868.2008.00700.x
- Wood, S.N. (2017) Generalized Additive Models: An Introduction with R (2nd edition). Chapman and Hall/CRC
- Zuur, A. F., Ieno, E. I. (2018). Beginner s Guide to Spatial, Temporal and Spatial-Temporal Ecological Data Analysis with R-INLA Volume II: GAM and zero-inflated models Published by Highland Statistics Ltd. Highland Statistics Ltd. Newburgh United Kingdom


## Exploratory {.tabset}

The data set contains the year of maturity, the month, the length (lt), the sex, the year of sample and the laboratory (institute) as you can see below. Note that for this study we have considered a subset of the data considering only females (sex=2).

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Load pckgs -------------------------------------------------------------------
library(mgcv)
library(INLA)
library(plyr)
library(tidyr)
library(GGally)
library(lattice)
library(ggplot2)
library(cowplot)
library(glmmTMB)
library(INLAutils)
library(gridExtra)

source("HighstatLibV11.R")

# Read data --------------------------------------------------------------------

rm(list=ls())
data <- read.csv("mat_data.csv", header = T, check.names = FALSE, 
                 sep = ";" ,dec = ".", stringsAsFactors = F)
data$year_mat <- as.factor(data$year_mat)
data$sex <- as.factor(data$sex)
data$lab <- as.factor(data$lab)
data$month <- as.factor(data$month)

# Subset females
last=unique(data$year)
data=subset(data,data$sex==2)
head(data)

# NA's in mat

ind=is.na(data$mat)
ind=which(ind==TRUE)
data=data[-ind,]
```


The following plot report the number of samples for each year and institute. IPMA has no maturity data for the following years: 1980-1991, 1995, 1996, 1999, 2000, 2017-2019. IEO data is provided for the completed time period 1980-2019. Note that 2020 maturity data was provided only in May by the IEO. Since the information for this year is incomplete and may cause bias in the estimation of the ogive it has been decided to eliminate it.

```{r echo=FALSE, message=TRUE, warning=FALSE, paged.print=FALSE}

# Exploratory plots ------------------------------------------------------------------------

## Number of samples by year/lab -----------------------------------------------
sample_yr <- as.data.frame(table(data$lab,data$year_mat))
names(sample_yr) <- c("lab","year","freq")
f1=ggplot(sample_yr, aes(x=year, y = freq, fill=lab)) + ggtitle("Number of samples by year/lab") +
   geom_bar(stat="identity",width=.7, position = "dodge") +theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
   theme(plot.title = element_text(size=10))



f1
```

Next plot reports the number of samples by month and institute. Maturity data was compiled from the IEO and IPMA samples only for the spawning season, December to May. Note that, samples collected in December were allocated to the following year. Larger IPMA sampling corresponds to February and March, whereas for the IEO the larger sampling corresponds to March and April.

```{r echo=FALSE, message=TRUE, warning=FALSE, paged.print=FALSE}


## Number of samples by month/lab -----------------------------------------------
sample_m <- as.data.frame(table(data$lab,data$month))
names(sample_m) <- c("lab","month","freq")
sample_m$month<- factor(sample_m$month, c("12", "1", "2", "3", "4", "5","6","7","8","9","10","11"))
f2=ggplot(sample_m, aes(x=month, y = freq, fill=lab)) + ggtitle("Number of samples by month/lab") +
   geom_bar(stat="identity",width=.7, position = "dodge") + theme(plot.title = element_text(size=10))


f2
```

Next plot reports the number of samples by length and institute (laboratory). Overall good sampling of relevant length classes (from 20cm to 70cm).

```{r echo=FALSE, message=TRUE, warning=FALSE, paged.print=FALSE}
## Number of samples by length -------------------------------------------------

cutoff_lengths <- c(seq(min(data$lt),19,by=1),seq(from=20, to=40, by=1), seq(from=42, to=70, by=2),seq(71,max(data$lt),by=4))
data$bin <- cut(data$lt, cutoff_lengths, labels = cutoff_lengths[-1])
table(data$lab,data$bin)


sample_lt <- as.data.frame(table(data$lab,data$bin))
names(sample_lt ) <- c("lab","length","freq")
f3=ggplot(sample_lt , aes(x=length, y = freq, fill=lab)) + ggtitle("Number of samples by length/lab") +
   geom_bar(stat="identity",width=.7, position = "dodge") + theme(plot.title = element_text(size=10),axis.text.x = element_text(angle=90, hjust=1))


f3
```

Following 2010 benchmark it was decided to cut the ogive assigning zero to lengths below 21 cm because they are not mature.

```{r echo=FALSE, message=TRUE, warning=FALSE, paged.print=FALSE}

## Values less that 21cm -------------------------------------------------------

aux=subset(data,data$lt<21)[,c(3,5)]
sum(aux$mat)
ind=which(aux$mat==1)
aux[ind,]

data$mat[data$lt < 21 ] <- 0
```


Next plot reports the number of samples by year, month and institute. The plot shows that previously to 2001 IPMA information is missing except for 1992, 1993, 1994, 1997 and 1998. Furthermore, IEO sample size before 2001 is low and for some years not all months of the spawning season has been sampled. According to that years 1980-2000 are grouped for the modeling. On the other hand, for years 2017-2019 there are not IPMA information and the IEO samples sizes are again low. Hence, such years are also grouped in the modeling.


**Hence, our year covariable is not the year specific level factor is a year specific category factor with the following categories: 1980-2000, 2001, 2002, 2003, 2004, 2005, 2006, 2007, 2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017-2019.**


```{r echo=FALSE, message=TRUE, warning=FALSE, paged.print=FALSE}
sample_yr_m <- as.data.frame(table(data$month,data$year_mat,data$lab))
names(sample_yr_m) <- c("month","year","lab","freq")
sample_yr_m$month<- factor(sample_yr_m$month, c("12", "1", "2", "3", "4", "5","6","7","8","9","10","11"))

ggplot(sample_yr_m, aes(x=year, y = freq, fill=month)) + ggtitle("Number of samples by year/month(Dec-May)") +
  geom_bar(stat="identity",width=.7, position = "dodge") +theme(axis.text.x = element_text(angle = 90, hjust = 1)) +  theme(plot.title = element_text(size=10)) + facet_wrap(~lab,nrow = 2)
```


## Motivation {.tabset}

The maturity data is provided by two countries, Portugal and Spain, and a  combined maturity ogive is required. Previous analysis provides evidences that in Portugal the maturity occurs at lower lengths than in Spain. In fact the regression logistic model (generalized linear model) below  explains the maturity (binary response, immature/mature) using the length and the country factor leading to two statistical different ogives for each country.

The maturity data covers from 1980 to 2019, however, while the Spanish data cover the entire period, we have missing Portugal data for some years, and furthermore the samples sizes by year for each country are not balanced. For that reason the unification of the maturity data on an unique sample ignoring the country for further modeling, using for example glm, is not a suitable option. Other option can be a weighted average of the country ogives, but for that it is necessary to decide which weights must be used. After some research, we have found a possible solution using a Bayesian approach.

Our proposal is a bivariate bayesian regression model using the integrated nested Laplace approximation (INLA) (Rue et al., 2009) approach in the R-INLA software (https://www.r-inla.org/). 

```{r}
df2 <- data
mod.lab2 <- glm(mat ~ lt*lab, family = binomial(logit), data = df2) 
summary(mod.lab2)

#L50 Females IEO
-(coef(mod.lab2)[1]/coef(mod.lab2)[2])
#L50 Females IPMA
-(coef(mod.lab2)[1]+coef(mod.lab2)[3])/(coef(mod.lab2)[2]+coef(mod.lab2)[4])
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
new.data <- expand.grid(lt= c(0:max(df2$lt)), 
                        lab = c(levels(df2$lab)))
preds <- predict(mod.lab2, newdata = new.data, type = 'link',se.fit = TRUE)
new.data$pred.full1 <- preds$fit
new.data$ymin1 <- new.data$pred.full1 - 1.96*preds$se.fit 
new.data$ymax1 <- new.data$pred.full1 + 1.96*preds$se.fit

new.data$pred.full <-mod.lab2$family$linkinv(new.data$pred.full1)
new.data$ymin <- mod.lab2$family$linkinv(new.data$ymin1)
new.data$ymax <- mod.lab2$family$linkinv(new.data$ymax1)

g2 <- ggplot(df2,aes(x = lt, y = mat, col=lab)) + 
 # facet_wrap(~lab) + 
  geom_point() + xlim(0, 115) + 
  geom_ribbon(data = new.data,aes(y = pred.full, ymin = ymin, ymax = ymax),alpha = 0.25) +
  geom_line(data = new.data,aes(y = pred.full)) + ggtitle ("females with 95% CI") + theme(plot.title = element_text(size=10))+
  annotate(geom = "text", x = 60, y = 0.5, label ="atop(L50[ieo] == 43.6,L50[ipma] == 41.7)" ,parse = TRUE)
g2

```


## Prepare data {.tabset}

The bivariate model response considers separetely two maturity variables one for each country. The two response variables are explained using length and year covariables. The model formulation in terms of covariables depends on the aim: 
- (i) a standard year combined maturity ogive or 
- (ii) a combined maturity ogive by year. 

On (i) the common predictor for the two responses is equal to an intercept plus a linear effect of the length plus a year random effect. The year random effect is changed by a year factor for (ii) approach.  The model carried out a combined estimation of all the parameters of the common predictor providing a combined maturity to introduce in the stock assessment model. 

**NOTE:** as mentioned previously year covariable has the following categories: 1980-2000, 2001, 2002, 2003, 2004, 2005, 2006, 2007, 2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017-2019.

```{r message=FALSE, warning=FALSE, paged.print=TRUE}
# Prepare data -----------------------------------------------------------------


NLbins<-c(seq(from=20, to=40, by=1),seq(from=42, to=70, by=2)) # Desired bins (SS model) 67
l_b=length(NLbins)

len=data$lt
l_len=length(len);aux=rep(0,l_len)

years<-(min(as.numeric(as.character(data$year_mat))):max(as.numeric(as.character(data$year_mat))))

# Response ----------------------------------------------------------------------

data_ieo=subset(data,data$lab=="ieo")
data_ipma=subset(data,data$lab=="ipma")
data=rbind(data_ieo,data_ipma)

ind_ieo=which(data$lab=="ieo")
ind_ipma=which(data$lab=="ipma")
len=length(data$lab)

len_ieo=length(ind_ieo)
len_ipma=length(ind_ipma)

YCombined <- matrix(NA, nrow = len, ncol = 2)
YCombined[1:len_ieo, 1]     <- (data$mat[ind_ieo])
YCombined[(len_ieo+1):(len_ipma+len_ieo), 2] <- (data$mat[ind_ipma])

# Grouped years ----------------------------------------------------------------

# Years previous to 2001 into a group ------------------------------------------

data$Gyear_mat=as.character(data$year_mat)
ind=which(as.numeric(as.character(data$year_mat))<2001)
data$Gyear_mat[ind]="1980_2000"

# Years 2017,2018 and 2019 into a group ----------------------------------------
ind=which(as.numeric(as.character(data$year_mat))>2016)
data$Gyear_mat[ind]="2017-2019"
   
data$Gyear_mat=as.factor(data$Gyear_mat)

```

## Model total  {.tabset}

**Standard ogive**: a single ogive for both institutes and years.

### Code

```{r message=FALSE, warning=FALSE, paged.print=TRUE}
# Model 1 ----------------------------------------------------------------------

f3 <-  YCombined ~ 1 + lt +  
                             f(Gyear_mat, model = "iid")

I3 <- inla(f3,
           control.compute = list(config=TRUE,
                                  dic = TRUE, 
                                  cpo=TRUE),
           family = c("binomial","binomial"),
           data = data, 
           control.inla = list(strategy = 'adaptive'), 
           verbose=TRUE, num.threads = 1)

summary(I3)
#INLAutils::plot_fixed_marginals(I3)
#INLAutils::plot_hyper_marginals(I3)
#INLAutils::plot_random_effects(I3)

# Prediction IPS ---------------------------------------------------------------
I1=I3
r=I3
r.samples = inla.posterior.sample(1000, r)
psam <- sapply(r.samples, function(x) {
   
   lt_effect <- x$latent %>% rownames(.) %>% stringr::str_detect("^lt") %>% x$latent[.,]
   intercept <- x$latent %>% rownames(.) %>% stringr::str_detect("^\\(Intercept\\)") %>% x$latent[.,]
   year_effect <- rnorm(length(lt_effect), sd = 1/sqrt(x$hyperpar[1]))
   predictor <- intercept + year_effect + lt_effect*NLbins
   exp(predictor)/(1 + exp(predictor))
})
q.sam_al_a <- apply(psam, 1, quantile,
                    c(.025, 0.05, 0.5, 0.95, .975), na.rm =TRUE)

```

### Plot

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
# Plot -------------------------------------------------------------------------

# Year iid
library(ggplot2)
suabinm <- I3$summary.random$Gyear_mat$mean
suabin2 <- I3$summary.random$Gyear_mat$`0.025quant`
suabin9 <-I3$summary.random$Gyear_mat$`0.975quant`
suabinID<-I3$summary.random$Gyear_mat$ID
suabin<-data.frame(suabinm, suabin2,suabin9,suabinID)

p1<-ggplot(data = suabin, aes(x = suabinID, y = suabinm, group=1))+
   geom_line(aes(x = suabinID, y = suabinm), color="#29AF7FFF", size=0.9)+ 
   geom_ribbon(aes(x = suabinID, ymin = (suabin2), ymax = (suabin9)), 
               alpha = 0.25, fill="gray70", linetype=1)+
   ggtitle(" ")+
   xlab("Year")+
   ylab("Year effect ")+
   ggtitle("Year random effect (iid)") +
   theme_light() +
   theme(axis.text.x = element_text(angle = 90))
p1

# Ogive
df_plot <- data.frame(t(q.sam_al_a))
rownames(df_plot) <- 1:(length(NLbins))
df_plot <- cbind(df_plot, 
                 lt = NLbins)

p2<-ggplot(df_plot) +
   geom_line(aes(x = lt, y = X50.,group=1), color="dodgerblue4", size=0.8) +
   geom_ribbon(aes(x = lt, ymin = (X2.5.), ymax = (X97.5.)), 
               alpha = 0.25, fill="gray70", linetype=1)+
   geom_line(aes(x = lt, (X2.5.)), color = "grey50", size = 0.1, linetype="dashed") + 
   geom_line(aes(x = lt, (X97.5.)), color = "grey50", size = 0.1, linetype="dashed") +
   xlab("Length (cm)") + ylab("Prob. Maturity") +
   theme_light() + xlim(10,80) + ggtitle("Standard maturity ogive")
  #geom_vline(xintercept=42.01015, 
  #              color = "red", size=0.5)+
  # geom_hline(yintercept=0.5, 
  #    color = "red", size=0.5)

p2

```


### $L_{50}$ values

Length at 50% maturity.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# L50 values -------------------------------------------------------------------

L50=-I3$summary.fixed[1,1]/I3$summary.fixed[2,1]


upper=-I3$summary.fixed[1,5]/I3$summary.fixed[2,5]

lower=-I3$summary.fixed[1,3]/I3$summary.fixed[2,3]


dL50=data.frame(L50,upper,lower)
colnames(dL50)=c("L50","lower","upper")
dL50
```


## Model by year {.tabset}

**Yearly ogive**: a specific ogive for year category.

### Code

```{r message=FALSE, warning=FALSE, paged.print=TRUE}
# Model 2 ----------------------------------------------------------------------

f3 <-  YCombined ~ 1 + lt +  Gyear_mat

I3 <- inla(f3,
           control.compute = list(config=TRUE,
                                  dic = TRUE, 
                                  cpo=TRUE),
           family = c("binomial","binomial"),
           data = data, 
           control.inla = list(strategy = 'adaptive'), 
           verbose=TRUE, num.threads = 1)
summary(I3)

# Prediction IPS ---------------------------------------------------------------
I2=I3
r=I3
r.samples = inla.posterior.sample(1000, r)
psam <- sapply(r.samples, function(x) {
   
   lt_effect <- x$latent %>% rownames(.) %>% stringr::str_detect("^lt") %>% x$latent[.,]
   intercept <- x$latent %>% rownames(.) %>% stringr::str_detect("^\\(Intercept\\)") %>% x$latent[.,]
   beta_y <- x$latent %>% rownames(. ) %>% stringr::str_detect("^Gyear_mat") %>% x$latent[.,]
   
   
   
   predictor1990 <- intercept + lt_effect*NLbins
   
   pre=list();l=length(beta_y)
   for (i in 1:l){
      pre[[i]]=intercept + beta_y[i] + lt_effect*NLbins
   }
   
   predictor=predictor1990
   
   for (i in 1:l){
      predictor <- c(predictor, pre[[i]])
   }
   
   exp(predictor)/(1 + exp(predictor))
})


q.sam_al_a <- apply(psam, 1, quantile,
                    c(.025, 0.05, 0.5, 0.95, .975), na.rm =TRUE)

```

### Plot

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
# Plot -------------------------------------------------------------------------
Gyears=unique(data$Gyear_mat)
df_plot <- data.frame(t(q.sam_al_a))
rownames(df_plot) <- 1:((length(NLbins))*length(Gyears))
df_plot <- cbind(df_plot, 
                 year = sort(rep(unique(data$Gyear_mat), (length(NLbins)))),
                 lt = (round(rep(NLbins,length(Gyears)))))

f5=ggplot(df_plot) +
   geom_ribbon(aes(x = lt, ymin = (X2.5.), ymax = (X97.5.), group=year), alpha = 0.25, fill="gray70", linetype=1)+
   geom_line(aes(x = lt, y = X50., group = year, col = year)) +xlab("") + ylab("") +
   xlab("Length (cm)") + ylab("Prob. Maturity") +
   ggtitle("Predicted ogives by year") +
   theme_light()
f5

# All years
ggplot(df_plot) +
      geom_line(aes(x = lt, y = X50.,group=1), color="dodgerblue4", size=0.7) +
      geom_ribbon(aes(x = lt, ymin = (X2.5.), ymax = (X97.5.)), 
               alpha = 0.25, fill="gray50", linetype=1)+
      #geom_line(aes(x = lt, (X2.5.)), color = "grey50", size = 0.1, linetype="dashed") + 
      #geom_line(aes(x = lt, (X97.5.)), color = "grey50", size = 0.1, linetype="dashed") +
      xlab("Length (cm)") + ylab("Prob. Maturity") +
      ggtitle("Predicted ogives by year")+
      theme_light() + facet_wrap(~year)




output_ogive_by_year=as.data.frame(cbind(df_plot$lt,df_plot$X50.,df_plot$X2.5.,df_plot$X97.5.))
output_ogive_by_year$year=df_plot$year
colnames(output_ogive_by_year)=c("length","mat","lower","upper","year")

```


### $L_{50}$

$L_{50}$ (length at 50% maturity) times series. Since the analysis of the series shows clear variability among year categories, the time specific model is proposed to be used instead to the standard year combined maturity ogive.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
dL50=data.frame(matrix(NA, nrow = length(Gyears), ncol = 3))

L50=-I3$summary.fixed[1,1]/I3$summary.fixed[2,1]


upper=-I3$summary.fixed[1,5]/I3$summary.fixed[2,5]

lower=-I3$summary.fixed[1,3]/I3$summary.fixed[2,3]

dL50[1,]=c(L50,upper,lower)

for (i in 2:length(Gyears)){
L50=-(I3$summary.fixed[1,1]+I3$summary.fixed[1+(i),1])/I3$summary.fixed[2,1]


upper=-(I3$summary.fixed[1,5]+I3$summary.fixed[1+i,5])/I3$summary.fixed[2,5]

lower=-(I3$summary.fixed[1,3]+I3$summary.fixed[1+i,3])/I3$summary.fixed[2,3]

dL50[i,]=c(L50,upper,lower)

}

colnames(dL50)=c("L50","lower","upper")
dL50$year=Gyears

dL50

# Plot L50 ---------------------------------------------------------------------


f6<-ggplot(dL50) +
   geom_line(aes(x = year, y = L50,group=1), color="dodgerblue4", size=0.8) +
   geom_ribbon(aes(x = year, ymin = (lower), ymax = (upper),group=1), 
               alpha = 0.25, fill="gray70", linetype=1)+
   geom_line(aes(x = year, (lower)), color = "grey50", size = 0.1, linetype="dashed") + 
   geom_line(aes(x = year, (upper)), color = "grey50", size = 0.1, linetype="dashed") +
   xlab("Year") + ylab("L50") +
   theme_light() +ggtitle("L50 by year")+
   theme( axis.text.x = element_text(angle=90, hjust=1))


f6
```



## Supplementary material {.tabset}

Structural changes

A structural change analysis has been applied over the year time series of $L_{50}$ (derived from the model using year factor covariable with a specific level for each year instead of the year categories). As you can see this analysis also reports 2000 as a break point of the time series in accordance with our conclusion after the exploratory analysis.

```{r message=FALSE, warning=FALSE, paged.print=TRUE}
library(strucchange)


load("50.RData")

maturity<-dL50

# Input NA's (if is required) 

interpFun <- function(dat) {
for (i in 1:length(dat)){
  if (is.na(dat[i])) 
    if(i == 1) {
      dat[i] <- rnorm(1,mean(dat, na.rm=T),
                       sd(dat, na.rm=T))
    } else {
      dat[i] <- rnorm(1,mean(dat[c(i-1, i+1)],na.rm=T),
                          sd(dat[c(i-1, i+1)],na.rm=T))
    }
}
return(dat)
}

# Define time series -----------------------------------------------------------
mInterp <- interpFun(maturity$L50)
mInterp <- ts(mInterp, 
                  start=min(maturity$year),
                  frequency = 1)

# Detect break and test-------------------------------------------------------- 
ocusm <- efp(mInterp~1, type="OLS-CUSUM") 
#ocusm <- efp(mInterp~1, type="Rec-CUSUM") 
#ocusm <- efp(mInterp~1, type="Rec-MOSUM") 
#ocusm <- efp(mInterp~1, type="OLS-MOSUM") 
bpm <- breakpoints(mInterp~1)
maturity$year[bpm$breakpoints]

sctest(ocusm)

# Plot series + break point ----------------------------------------------------


plot(mInterp,
     xlab= "Year",
     ylab= "L50 parameter",
     lty=1,
     lwd=2,
     main = "L50 breakpoints")
lines(mInterp, 
      lty = 1, 
      lwd = 2)
abline(v=maturity$year[bpm$breakpoints],
       lwd= 2,
       lty = 1,
       col="blue")
legend("topright",
       legend = c("a", "bp L50"),
       lty = c(1,1),
       col = c("black", "blue"), bty="n", x.intersp=0.5, horiz= F, cex=0.70)


```

